---
title: "Importar os dados"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
BiocManager::install("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
```

```{r,message=FALSE,warning=FALSE}
library(GEOquery)
library(minfi)
library(IlluminaHumanMethylationEPICmanifest)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
```

Dados disponíveis em: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE168739


The raw files are deposited in GEO and can be downloaded by using the getGEOSuppFiles. There are two files for each replicate, since the EPIC array is a two-color array. The files are downloaded in compressed format and need to be uncompressed before they are read by the `read.metharray.exp` function.

Para darmos início à análise dos dados, vamos obter os dados sem qualquer pré-processamento:

```{r,eval=FALSE}
getGEOSuppFiles("GSE168739")
untar("GSE168739/GSE168739_RAW.tar", exdir = "GSE168739/idat")
list.files("GSE168739/idat", pattern = "idat")
files <- list.files("GSE168739/idat", pattern = "idat.gz$", full = TRUE)
sapply(files, gunzip, overwrite = TRUE)
```


Agora os dados podem ser importados no R:

```{r,eval=FALSE}
rgSet <- read.metharray.exp("GSE168739/idat",force=TRUE)
dim(rgSet)
```
To access phenotype data we use the `pData` function. The phenotype data is not yet available from the `rgSet`.


```{r,eval=FALSE}
pData(rgSet)
```

```{r,eval=FALSE}
geoMat <- getGEO("GSE168739")
pD.all <- pData(geoMat[[1]])

pD <- pD.all[, c("geo_accession", "gender:ch1",
                 "age:ch1")]
```

This phenotype data needs to be merged into the methylation data. The following commands guarantee we have the same replicate identifier in both datasets before merging.

```{r,eval=FALSE}
sampleNames(rgSet) <- sapply(sampleNames(rgSet),function(x)
  strsplit(x,"_")[[1]][1])
rownames(pD) <- pD$geo_accession
pD <- pD[sampleNames(rgSet),]
pData(rgSet) <- as(pD,"DataFrame")
save(rgSet,file="../Analises/resultados/rgSet.rds")
```


```{r,eval=FALSE}
load("../Analises/resultados/rgSet.rds")
dim(rgSet)
n = dim(rgSet)[2]

# work with a small sample 

set.seed(12021)
samp <- sample(1:n,size=60,replace=FALSE)

rgSet <- rgSet[,samp]
phenoData <- pData(rgSet)

save(rgSet,file="../Analises/resultados/rgSetSample.rds")
```


```{r}
load("rgSetSample.rds")
densityPlot(rgSet)
```

```{r}
dim(rgSet)
```

```{r}
mSetSq <- preprocessQuantile(rgSet)
```

```{r,fig.width=15}
par(mfrow=c(2,1))
densityPlot(rgSet,main="Raw", legend=FALSE)
densityPlot(getBeta(mSetSq),main="Normalized", legend=FALSE)
```


```{r, fig.width=15}
par(mfrow=c(1,2))
boxplot(getBeta(rgSet))
boxplot(getBeta(mSetSq))
```

































